Parameter-less Population Pyramid

### 2014-01-27 Parameterless Population Pyramid/main.tex
Minor tweaks.

The problem of parameter tuning has encouraged work to remove parameters from
user control in an effort to improve out-of-the-box GA quality. In
an early example, a crossover based GA was devised which used
concurrent racing populations which removed all user specified
parameters~\cite{harik:1999:parameterlessga}.
Population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}.
Without the need to set a population size, racing ensures the GA will take no more than
a logarithmic factor increase in evaluations beyond optimal.
Some experimental evidence suggests the bound is closer to a constant
factor worse than optimal.

@@

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many EC techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optima using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to continue,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

@@

\subsection{Pyramid}
\label{sec-pyramid}
As mentioned, P3 maintains a pyramid like structure of populations
with each level representing something akin to different generations
of evolution. More specifically, a solution's level is one higher than
its highest parent, similar to the layering done
in~\cite{hornby:2006:alps}.

@@

\subsection{Crossover}
\label{sec-crossover}
P3 uses a crossover method derived from that defined for LTGA~\cite{thierens:2013:ltgahiff}
(Section~\ref{sec-ltga});
P3 uses gene value entropy to construct clusters of genes that should have their values
conserved during crossover.  There are two primary steps in
defining this type of crossover: \emph{cluster creation} and \emph{cluster usage}.

@@

P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution.  For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in that cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through fitness plateaus.

@@

To complete a single crossover event, all of the useful clusters are tested, each
requiring at most a single evaluation.  Clusters are tested in smallest first order,
with clusters of equal size ordered randomly.  This ensures smaller building blocks
are optimized before large changes are made to the genome, thus helping preserve
diversity.  This also helps prevent genetic hitchhikers, which may interfere with
entropy calculations.  As cluster building and application ordering are not dependent on
the position of loci in the genome, this operator introduces no positional bias in the search points
evaluated.

@@

P3 makes very weak
assumptions about the search landscape.  In many evolutionary
systems the crossover operator assumes linked genes
are collocated in the genome.  This requires the user
to configure genome structure based on gene linkage information they
are unlikely to have. All of
P3's operations ignore gene ordering, removing that level of
configuration from the user.

@@

Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
This means each time a solution is added to $P_0$, at least $O(N)$ evaluations
must be performed.  This implies that the $O(N^2)$ cost of the update to $P_0$ can
be amortized over these evaluations, making this update $O(N)$ per evaluation.
%Note
%that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
%individual is already at a local optimum.  Furthermore, only restarts that result
%in a unique individual cause updates to $P_0$.

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied, each causing their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $O(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost to update $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a cluster.  As such we are currently unable to state conclusively
that updating $P_i$ will require $O(N)$ operations per evaluation.  That said,
empirically it appears to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

There is one more operation in P3 that may not scale linearly with
genome size per evaluation: finding a non-identical donor.  When
performing crossover P3 iteratively searches a population for a random
donor which contains genes which differ from the current solution for
the current crossover cluster.  In the worst case, this operation
could require $|C_j||P_i|$, where $|C_j|$ is the number of genes in
the crossover and $|P_i|$ is the number of solutions in the
population.  However, $|P_i|$ grows very slowly.  Also, in empirical
evaluation, P3's wall clock time per evaluation appears to scale
linearly with problem size.

@@

\subsection{Linkage Tree Genetic Algorithm}
\label{sec-ltga}
The closest relative to P3 in existing literature is the Linkage Tree Genetic Algorithm
(LTGA).  There have been a number of different variations proposed to LTGA since
its original publication~\cite{thierens:2010:ltga}, so for comparison we have chosen
to use the variant described in the most recent publication~\cite{thierens:2013:ltgahiff}.
This algorithm is also the current state of the art, outperforming other algorithms
in black box optimization across numerous benchmarks.

@@

To ensure maximal LTGA quality, we use the bisection method~\cite{goldman:2012:ltga}
to minimally set LTGA's population size.  To be considered successful, a
population size must result in 100 runs without premature convergence.
According to the Rule of Three~\cite{jovanovic:1997:ruleofthree}, this bounds
LTGA's failure rate to less than $\frac{3}{100+1} \approx 3\%$ with 95\% confidence.
Note that runs used in setting the population are considered testing and were not
used during analysis.  On randomized problems, training was performed on a disjoint
set of problem instances.

@@

Other variants of LTGA have more closely resembled P3~\cite{goldman:2012:ltga},
but as those variants are not in active use, we did not use them for comparison.
However, no LTGA variant has removed the population size parameter
or fully addressed the problem of premature convergence.

%TODO Consider removing this
There have been other model based optimization techniques that use little to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
%In general, LTGA is the most powerful general optimizer with few parameters.

@@

We also made minor modifications to selection and to prevent wasting evaluations.
If the best offspring produced in a generation is a mutant offspring, select that
over any of those produced using crossover.  If there is a fitness tie between the
best offspring in a generation, select the one with the maximum hamming distance
to the parent.  This encourages drift over plateaus.  While the original paper
discusses a ``mod'' version of the $\lambda$ control strategy for how to handle
when offspring of equal fitness are produced, we found this conflicted with our
method of restarting.  As such we use the original control strategy. %, where $\lambda$
%is increased if the best offspring in a generation is not strictly better than its parent.
If an offspring produced by crossover has identical genes to either of its parents,
it is not evaluated.

@@

\subsection{Randomly Generated Problem Classes}
By enforcing only certain characteristics of a search space while allowing others
to be generated randomly, we can perform tests on classes of problems instead
of specific instances.  This can improve the generality of results as experiments
then predict the expected quality over the entire class.  Unlike single
instance problems, the challenge with randomly generated problems is often how
to know when the global optimum is reached.

@@

Our final class is a subset of the Maximum Satisfiability (MAX-SAT) problem.
This problem consists of a series of disjunctive clauses, such that each clause contains
three problem variables, some of which are negated. While
MAX-SAT requires exponential time to solve, we have devised a method of generating
only satisfiable instances.
First, a random target solution is generated.  For each clause, set the signs of each
term ensuring the target solution satisfies that clause.
To ensure there is no bias to term signs in the generated problem there is a $\frac{1}{6}$ probability
all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
and a $\frac{4}{6}$ probability only one sign matches.  Note that its possible for there to be
other solutions that also satisfy all clauses.  We chose a clause to variable ratio
of $r=4.27$.

\section{Experimental Results}
\begin{figure*}
  \centering
  \includegraphicswide{WithRast}
  \caption{Comparison of the median number of evaluations to reach the global optimum for
           the four different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no point
           is drawn.  Results given on a log-log scale.}
  \label{fig-results}
\end{figure*}

@@

On all 7 problems, the First Improvement Hill Climber and the $(1+(\lambda,\lambda))$
optimizers appear to have a higher growth complexity than P3.  On all of the problems
except MAX-SAT they are also worse than LTGA.  This makes sense as all of the
benchmarks are deceptive in some sense, and neither of these techniques is designed
to handle deceptive landscapes.  The inability for the iterative hill climber to solve
any of the problems indicates that P3's success is more than just performing local search.

Figure~\ref{fig-results} does provide one misleading
exception for $(1+(\lambda,\lambda))$ on the Deceptive Trap problem, where it appears to
level out at high genome sizes.  This is entirely an artifact of the extremely high
variance of this algorithm on this problem.  Note that many of the large problem
sizes failed to find the optimum for the majority of runs (no dot drawn for that size).

The behavior of P3 and LTGA on the four single instance problems appears very similar,
with P3 generally being lower by a constant factor.  This makes sense as LTGA's
tuning allows it to determine a single population size sufficient to solve that
instance.  However, because it uses a fixed population size, it must set the that
size large enough to ensure all random initializations of its population
contain enough diversity to reach the global optimum.  By comparison, P3 is able to
iteratively increase the population size, stopping once enough diversity has been generated.
As such we would expect P3 to do better, but not a lot better.

@@

Table~\ref{table-results} provides a detailed look at the largest problem sizes
in which LTGA was able to find the global optimum reliably in less than 100 million
evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
in p-values less than $10^{-22}$, with the exception of Deceptive Step Trap which
resulted in $0.00025$.  As such these results are highly significant.  The amount
of speedup ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
cases P3 outperformed LTGA and all other comparison techniques.

As verification that our tuning for LTGA is valid, we refer
to~\cite{thierens:2013:ltgahiff}, which tested the same variant of LTGA that we
used.  In the section on Child Node filtering, they reported approximately 200,000
mean evaluations to success on the HIFF problem using $N=512$.  Our result of
239,365 is of comparable quality, with P3 solving the same problem using an average
of 92,129 evaluations.  The discrepancy is likely due to the cited work using
their tuning runs in their average, unrealistically allowing each run to use a different tuned population size.

@@

Using relaxed requirements, this can likely be extended to any arbitrary fully decomposable
problem. Assuming that for sufficiently large population sizes the entropy between
bits in a single subproblem is lower than the entropy of bits between subproblems,
P3 will construct a cluster capable of moving exactly one subproblem's gene values
during crossover.  As crossover cannot decrease the fitness of a solution each
crossover will have a greater than zero probability of improving subproblem solution quality.
The probability of low fitness local optima for a subproblem reaching high pyramid levels
is therefore low, which increases the probability of donors only having high quality
solutions to subproblems.  Combined this is expected to create a runtime potentially
exponential in subproblem size, but polynomial in problem size.

This proof sketch can also be extended to the HIFF problem, with minor modifications.
Each application of the hill climber ensures all subproblems of size 2 will be solved.
Once $P_0$ has a sufficient number of solutions it will correctly cross all subproblems of size 2
independently, such that all future solutions crossed with $P_0$ will have all subproblems
of size 4 solved.  While $P_1$ can contain solutions that do not correctly solve some 2 bit
subproblems (added before $P_0$ reached critical mass), the probability of these solutions
being selected as donors decreases rapidly (solutions added after $P_0$ reaches critical mass).
As such each level effectively serves to optimize the next largest subproblem size, leading to
convergence using a logarithmic number of levels.

\section{Conclusions and Future Work}
While many parameter-less techniques trade the potential quality of optimal tuning
for general effectiveness, P3 appears to robustly outperform all of the comparison
state of the art techniques across a diverse set of benchmark problems.  This is
because unlike other parameter-less techniques, P3 does not duplicate effort running
multiple populations, nor does it throw away learned information like many evolutionary
techniques.  By striking
a balance between continuously adding diversity and the extreme exploitation provided
by the multiple evaluation crossover, P3 is able to solve easy problems quickly
and hard problems eventually without any problem specific tuning.

Even when compared against LTGA, the closest relative to P3 in current literature
and current state of the art linkage learning algorithm, P3 surpasses expectation.
This is even after LTGA was extensively tuned to each problem class, an unrealistic
proposition on any real world problem.
%A likely advantage is that P3 is able to
%automatically scale to problem difficulty, adding diversity only as the problem
%requires.
%TODO Consider reiterating stuff about gene location invariant, scaling invariant.

P3's ability to efficiently optimize not just single problem instance classes,
but randomly generated problem classes without parameter setting makes it an
ideal candidate for real world application.  Its focus on exploiting all known
information before spending time increasing diversity makes it a great ``stop anytime''
algorithm.
%In future work, we plan to investigate in more detail P3's time to
%``good enough'' solutions, not just optimal.

@@

Currently, P3 is limited to single objective, noiseless, discrete valued problems.
In future work we plan to extend (or show no modification is necessary for) P3 to handle
problems from these classes.  Doing so will help make P3 more applicable to a wider
range of real world problems.

>>>>

The problem of parameter tuning has encouraged work to remove parameters from
user control in an effort to improve out-of-the-box GA quality. In
an early example, a crossover based GA was devised which used
concurrent racing populations which removed all user specified
parameters~\cite{harik:1999:parameterlessga}.
Population racing has also been paired with more advanced evolutionary
system, such as Hierarchical Bayesian
Optimization~\cite{pelikan:2004:parameterlesshboa}.
Without the need to set a population size, racing ensures the GA will take no more than
a logarithmic factor increase in evaluations beyond optimal.
While experimental evidence suggests racing is only a constant
factor loss, this method explicitly trades efficiency for applicability.

@@

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many EC techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optima using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to perform crossover,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

@@

\subsection{Pyramid}
\label{sec-pyramid}
As mentioned, P3 maintains a pyramid like structure of populations
with more optimized solutions in higher levels. More specifically, a solution's level is one higher than
its highest parent, similar to the layering done
in~\cite{hornby:2006:alps}.

@@

\subsection{Crossover}
\label{sec-crossover}
P3 uses a crossover method derived from that defined for LTGA~\cite{thierens:2013:ltgahiff}
(Section~\ref{sec-ltga});
P3 uses gene value entropy to construct clusters of genes that should have their values
conserved during crossover.  There are two steps in
defining this type of crossover: \emph{cluster creation} and \emph{cluster usage}.

@@

P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution as follows.  For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in that cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through fitness plateaus.

@@

To complete a single crossover event, all of the useful clusters are tested, each
requiring at most a single evaluation.  Clusters are tested in smallest first order,
with clusters of equal size ordered randomly.  This ensures smaller building blocks
are optimized before large changes are made to the genome, thus helping preserve
diversity.  This also helps prevent genetic hitchhikers, which may interfere with
entropy calculations.  As cluster building and application ordering are not dependent on
the position of loci in the genome, this operator introduces no positional bias on search.

@@

P3 makes very weak
assumptions about the search landscape.  In many evolutionary
systems the crossover operator assumes linked genes
are collocated in the genome.  This requires the user
to configure genome structure based on gene linkage information they
are unlikely to have. All of
P3's operations ignore gene ordering, allowing the user to ignore
that aspect of configuration.

@@

Each time a random restart occurs, at minimum $N+1$ evaluations are performed.
This means each time a solution is added to $P_0$, at least $\Omega(N)$ evaluations
must be performed.  This implies that the $O(N^2)$ cost of the update to $P_0$ can
be amortized over these evaluations, making this update $O(N)$ per evaluation.
%Note
%that $N+1$ is a very weak lower bound, as this only occurs if the randomly generated
%individual is already at a local optimum.  Furthermore, only restarts that result
%in a unique individual cause updates to $P_0$.

To cause an update to $P_i$, all of the crossovers discovered for
$P_{i-1}$ must first be applied, each causing their own evaluation.
If P3 were to na\"ively use all crossovers and perform an evaluation
each time, this would cause $\Omega(N)$ evaluations.  Again, this would
result in $O(N)$ amortized cost to update $P_i$.  However,
Section~\ref{sec-crossover} discussed how P3 will ignore some clusters
and will also skip an evaluation if no change to the solution was made
by a cluster.  As such we are currently unable to state conclusively
that updating $P_i$ will require $O(N)$ operations per evaluation.  That said,
empirically it appears to be true.  Furthermore, it seems counterintuitive
that intentionally wasting evaluations would be considered beneficial
to runtime.

There is one more operation in P3 that may not scale linearly with
genome size per evaluation: finding a non-identical donor.  When
performing crossover P3 iteratively searches a population for a random
donor with genes that differ from the current solution for
the current crossover cluster.  In the worst case, this operation
could require $|C_j||P_i|$, where $|C_j|$ is the number of genes in
the crossover and $|P_i|$ is the number of solutions in the
population.  However, $|P_i|$ grows very slowly.  Also, in empirical
evaluation, P3's wall clock time per evaluation appears to scale
linearly with problem size.

@@

\subsection{Linkage Tree Genetic Algorithm}
\label{sec-ltga}
The closest relative to P3 in the existing literature is the Linkage Tree Genetic Algorithm
(LTGA).  There have been a number of different variations proposed to LTGA since
its original publication~\cite{thierens:2010:ltga}, so for comparison we have chosen
to use the variant described in the most recent publication~\cite{thierens:2013:ltgahiff}.
This algorithm is also the current state of the art, outperforming other algorithms
in black box optimization across numerous benchmarks.

@@

To ensure LTGA is able to find the global optimum quickly, we use the bisection method~\cite{goldman:2012:ltga}
to minimally set LTGA's population size parameter.  To be considered successful, a
population size must result in 100 runs without premature convergence.
According to the Rule of Three~\cite{jovanovic:1997:ruleofthree}, this bounds
LTGA's failure rate to less than $\frac{3}{100+1} \approx 3\%$ with 95\% confidence.
Note that runs used in setting the population are considered testing and were not
used during analysis.  On randomized problems, training was performed on a disjoint
set of problem instances.

@@

Other variants of LTGA have more closely resembled P3~\cite{goldman:2012:ltga},
but as those variants do not appear to be in active use, we did not use them for comparison.
However, no LTGA variant has removed the population size parameter
or fully addressed the problem of premature convergence.
%TODO Consider removing this
There have also been other model based optimization techniques that use few to
no parameters.  Specifically LO-LIMD~\cite{posik:2011:parameterless} uses a
population size of one and a parameter free algorithm to perform optimization.
It relies on the existence of completely separable subproblems and noise free evaluation in order to determine
if two genes are linked through perturbation of their values.  While effective on
this class of problems, most real world problems do not fall into this category.
%In general, LTGA is the most powerful general optimizer with few parameters.

@@

We also made minor modifications to improve selection and to prevent wasting evaluations, as follows.
If the best offspring produced in a generation is a mutant offspring, select that
over any of those produced using crossover.  If there is a fitness tie between the
best offspring in a generation, select the one with the maximum hamming distance
to the parent.  This encourages drift over plateaus.  While the original paper
discusses a ``mod'' version of the $\lambda$ control strategy for how to handle
when offspring of equal fitness are produced, we found this conflicted with our
method of restarting.  As such we use the original control strategy. %, where $\lambda$
%is increased if the best offspring in a generation is not strictly better than its parent.
Finally, if an offspring produced by crossover has identical genes to either of its parents,
it is not evaluated.

@@

\subsection{Randomly Generated Problem Classes}
By enforcing only certain characteristics of a search space while allowing others
to be generated randomly, we can perform tests on \emph{classes} of problems instead
of specific instances.  This can improve the generality of results as experiments
then predict the expected quality over the entire class.  Unlike single
instance problems, the challenge with randomly generated problems is often how
to know when the global optimum is reached.

@@

Our final class is a subset of the Maximum Satisfiability (MAX-SAT) problem.
This problem consists of a series of disjunctive clauses, such that each clause contains
three problem variables, some of which are negated. While
MAX-SAT requires exponential time to solve, we have devised a method of generating
only satisfiable instances.
First, a random target solution is generated.  For each clause, set the signs of each
term ensuring the target solution satisfies that clause.
To ensure there is no bias to term signs in the generated problem, there is a $\frac{1}{6}$ probability
all clause signs match the target, a $\frac{1}{6}$ probability two of the three signs match,
and a $\frac{4}{6}$ probability only one sign matches.  Note that it's possible for there to be
other solutions that also satisfy all clauses.  We chose the standard clause to variable ratio
of $r=4.27$.

\section{Experimental Results}
\begin{figure*}
  \centering
  \includegraphicswide{WithRast}
  \caption{Comparison of the median number of evaluations to reach the global optimum for
           the four different optimization methods with respect
           to problem size.  If the median run did not reach the global optimum no data element
           is shown.  Results given on a log-log scale.}
  \label{fig-results}
\end{figure*}

@@

On all 7 problems, the First Improvement Hill Climber and the $(1+(\lambda,\lambda))$
optimizers appear to have a higher growth complexity than P3.  On all of the problems
except MAX-SAT they are also worse than LTGA.  This makes sense as all of the
benchmarks are deceptive in some sense, and neither of these techniques is designed
to handle deceptive landscapes.  The inability for the iterative hill climber to quickly solve
the problems indicates that P3's success is more than just performing local search.

Figure~\ref{fig-results} does provide one misleading
exception for $(1+(\lambda,\lambda))$ on the Deceptive Trap problem, where it appears to
level out at high genome sizes.  This is entirely an artifact of the extremely high
variance of this algorithm on this problem.  Note that many of the large problem
sizes failed to find the optimum for the majority of runs (no data element shown for that size).

The behavior of P3 and LTGA on the four single instance problems appears very similar,
with P3 generally being lower by a constant factor.  This makes sense as LTGA's
tuning allows it to determine a single population size sufficient to solve that
instance.  However, because it uses a fixed population size, it must set the that
size large enough to ensure all random initializations of its population
contain enough diversity to reach the global optimum.  By comparison, P3 is able to
iteratively increase the population size, stopping once enough diversity has been generated.
As such we would expect P3 to do better, but not a lot better than tuned LTGA.

@@

Table~\ref{table-results} provides a detailed look at the largest problem sizes
in which LTGA was able to find the global optimum reliably in less than 100 million
evaluations.  For each, we performed a pairwise Mann-Whitney~U test, all resulting
in p-values less than $10^{-22}$, with the exception of Deceptive Step Trap which
resulted in $0.00025$.  As such these results are highly significant.  The amount
of speedup as compared to optimally tuned LTGA ranged from 1.1 on Deceptive Step Trap to 942 on MAX-SAT, but in all
cases P3 outperformed LTGA and all other comparison techniques.

As verification that our tuning for LTGA is valid, we refer
to~\cite{thierens:2013:ltgahiff}, which tested the same variant of LTGA that we
used.  In the section on Child Node filtering, they reported approximately 200,000
mean evaluations to success on the HIFF problem using $N=512$.  Our result of
239,365 is of comparable quality, with P3 solving the same problem using an average
of 92,129 evaluations.  The discrepancy is likely due to the cited work using
their tuning runs in their average.

@@

Using relaxed requirements, this can likely be extended to any arbitrary, fully decomposable,
problem. Assuming that for sufficiently large population sizes the entropy between
bits in a single subproblem is lower than the entropy of bits between subproblems,
P3 will construct a cluster capable of moving exactly one subproblem's gene values
during crossover.  As crossover cannot decrease the fitness of a solution each
crossover will have a greater than zero probability of improving subproblem solution quality.
The probability of low fitness local optima for a subproblem reaching high pyramid levels
is therefore low, which increases the probability of donors only having high quality
solutions to subproblems.  Combined, this is expected to create a runtime potentially
exponential in subproblem size, but polynomial in problem size.

This proof sketch can also be extended to the HIFF problem with minor modifications.
Each application of the hill climber ensures all subproblems of size 2 will be solved.
Once $P_0$ has a sufficient number of solutions it will correctly cross all subproblems of size 2
independently, such that all future solutions crossed with $P_0$ will have all subproblems
of size 4 solved.  While $P_1$ can contain solutions that do not correctly solve some 2 bit
subproblems (added before $P_0$ reached critical mass), the probability of these solutions
being selected as donors decreases rapidly (solutions added after $P_0$ reaches critical mass).
As such each level effectively serves to optimize the next largest subproblem size, leading to
convergence using a logarithmic number of levels.

\section{Conclusions and Future Work}
While many parameter-less techniques trade the potential quality of optimal tuning
for general effectiveness, P3 appears to robustly outperform all of the comparison,
state of the art, techniques across a diverse set of benchmark problems.  This is
because, unlike other parameter-less techniques, P3 does not duplicate effort running
multiple populations, nor does it throw away learned information like many evolutionary
techniques.  By striking
a balance between continuously adding diversity and the extreme exploitation provided
by the multiple evaluation crossover, P3 is able to solve easy problems quickly
and hard problems eventually without any problem specific tuning.

Even when compared against LTGA, the closest relative to P3 in current literature,
P3 performs better.
This is even after LTGA was extensively tuned to each problem class, an unrealistic
proposition on any real world problem.
%A likely advantage is that P3 is able to
%automatically scale to problem difficulty, adding diversity only as the problem
%requires.
%TODO Consider reiterating stuff about gene location invariant, scaling invariant.

P3's ability to efficiently optimize not just single problem instance classes,
but randomly generated problem classes, without parameter tuning makes it an
ideal candidate for real world applications.  Its focus on exploiting all known
information before spending time increasing diversity makes it a great ``stop anytime''
algorithm.
%In future work, we plan to investigate in more detail P3's time to
%``good enough'' solutions, not just optimal.

@@

Currently, our results are limited to single objective, noiseless, discrete valued problems.
In future work we plan to extend (or show no modification is necessary for) P3 to handle
problems from these classes.  Doing so will help make P3 more applicable to a wider
range of real world problems.
